\chapter*{Acknowledgements}

First and foremost I would like to thank my supervisors Professor Mick Grierson and Professor Frederic Fol Leymarie for their dedicated support over the long journey of this research and my prior academic training, and to the wider community of academics in the department of Computing at Goldsmiths who supported me over my many years there. I will be eternally grateful to those at Goldsmiths who had the dedication and vision to establish Creative Computing as an academic discipline and create a place for people like me to thrive in. 

I would like to thank Professor Rebecca Fiebrink and Professor Matthew Yee-King for the invaluable feedback in my upgrade examination, which massively influenced the final shape of this thesis. Additionally, I would like to thank the examiners of this PhD, Professor Phillipe Pasquier and Professor Marco Gillies, for their time and the thoughtful questions they prepared for what was a very enjoyable viva. 

I would like to thank my friend and PhD colleague Sebastian Berns for coining the term \textit{active divergence} that succinctly defined and tied together what I had thought were very disparate pieces of research when I had originally conducted them. I would like to thank the principal investigators, admin team and the other fellow students from the Centre for Doctoral Training in Intelligent Games and Game Intelligence (IGGI; grant EP/L015846/1) for providing the support and community to make this PhD possible. I would also like to thank Sara Bodinar and Jan Birley from Writers Retreat UK for providing the space and mentorship during my retreat there in 2023 where the final framing and narrative of this thesis took shape. 

I would like to thank the curators and other organisers who chose to exhibit and showcase artworks made during this thesis, which include Luba Elliot, Xavier Snelgrove, Gemma Murray, Susanna Pousette, Paula Perissinotto, Clarissa Oliveira, Mathieu Arbez Hermoso and Rita Hajj. I would also like to thank Georgie Hoare, Joe Bedell-Brill from the band 0171 to use their commission as a test-bed for some of this research and document that process in this thesis.

I would like to thank my good friend Joe for his generosity and support during the first COVID lockdown and the support he gave to this research during that pivotal time. I would like to thank my partner Katie, for her patience and the resolute support that I so desperately needed in the final stages of this PhD. Finally, I would finally like to thank my Mum and Dad for the constant and unwaivering support they have given me throughout my academic journey, none of this would have been possible without them. 

